---
title: "Introduction to R Programming"
author: "By: Jasper Otieno"
date: "Last updated on 20/Sept/2023"
output:
  html_document: 
    toc: yes
    number_sections: yes
    theme: readable
    fig_caption: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---
_______________________________________________________________________________
                                                 
This is an introduction to R programming with end goal of learning the basics of R and finally being able to program in R and proceed to advanced programming.

# BASIC R PROGRAMMING

## DATA TYPES
R stores data in different types or classes. There are about 4 classes/types of data that R can store. To check the kind of class of data you are working with, use the function `class()` and use the function `typeof()` to check the type of data.

### Character
Refers to data in form of letters, sentences or numbers formatted to texts/strings e.g, names of people (James, rose), letter A, B, or colours, green, blue. They must always be wrapped in double quotation marks when referenced.

```{r}
(char1 <- c("ann", "erick", "", "l"))
class(char1)
typeof(char1)
```

### Numeric
Are class of data that are stored in number formats. These include whole numbers, decimals or even dates. There are various types of numeric data which include integers, double or just numeric.

```{r}
(numeric1 <- c(1, 2, 3, 4, 5, 6.5)) #double

class(numeric1)
typeof(numeric1)

```

### Integers
Are numeric data in form of whole numbers. They are denoted by letter L next to them.
```{r}
(numeric2 <- c(1L, 2L, 3L, 4L, 5L)) #integer (whole numbers)

class(numeric2)
typeof(numeric2)
```

### Factor
Refers to class of data that stores data in categorical format. They values take whole numbers hence are integer type.

```{r factors, include=FALSE}
(fact1 <- factor(c("ECDE", "Primary", "Secondary", "University"))) 

class(fact1)
typeof(fact1)

#Assigning values to factors
#fact2 <- as.factor(fact1, 
#                   levels = c("Primary", "Secondary", "University"),
#                   labels = c("Primary", "Secondary", "University"), 
#                   values = c(1, 2, 3)) #has an error. to investigate cause of #error. Investigate during data cleaning

#Converting factor to character
(fact3 <- factor(fact1))
```

### Logical
These are data types that take the values TRUE or FALSE. They are special kind of numeric data that indicates if a condition is true or not.

```{r}
(lg1 <- c(TRUE, FALSE, FALSE, TRUE))

class(lg1)
typeof(lg1)
```

### Dates and Times
Data can be stored in form of dates or times. They are special kind of numeric data of type double.

```{r}
#System Datetime
(date1 <- Sys.time())

class(date1)
typeof(date1)

#System date
(date2 <- Sys.Date())

class(date2)
typeof(date2)
```

### Data Type Conversion
You may want to convert one type of data to another. More often, converting numeric to character and vice versa. The conversion is largely depending on need of analysis or for storage.

```{r}
(numeric1 <- c(1, 2, 3, 4, 5, 6.5))
(numeric2 <- c(1L, 2L, 3L, 4L, 5L))

(data6 <- as.character(numeric1)) #converting numeric to character

class(data6)
typeof(data6)

(data7 <- as.numeric(data6)) #converting character to numeric

class(data7)
typeof(data7)

(data8 <- as.numeric(numeric2)) #converting integer to numeric

class(data8)
typeof(data8)
```

## DATA STRUCTURES
These are ways of organising data for ease of analysis or storage. There are 4 ways that data can be organised in R. To know the structure of your data, use the function `str()`.

### Vectors
Are the smallest or easiest way of organising data. They are as simple as just one character or letter. Vectors are single dimensional and homogeneous (only stores one type of data).

```{r}
#numeric vectors
(vect1 <- c(1, 3, 5, 68, 0)) #specified numbers, integers
(vect2 <- 1:10) #vector of sequential numbers 1 to 10. Default constant is 1
(vect4 <- 0:30) #vector of numbers increasing from 0 to 30. Reverse is true i.e., 30:0
(vect6 <- 2.5:7.5)

class(vect1)
typeof(vect1)
str(vect1)

#character vectors
(vect7 <- c("Jasper", "Dave", "Green", "A", ""))
str(vect7)

#Logical vectors
(vect8 <- c(TRUE, TRUE, FALSE, TRUE))
str(vect8)

#Factor vector
(gender <- c("F", "M", "F", "M", "M"))

#convert gender to a factor
(gender1 <- factor(gender))

(fcoulor <- factor(c("blue", "green", "red", "pink", "green")))

```

#### Forming sequence of numbers
```{r}
(vect9 <- seq(from = 1, 
             to = 25, 
             by = 2)) #creates sequence of numbers between 1 and 25 increasing by 2

(vect10 <- seq(from = 1, 
             to = 25, 
             length.out = 50)) # creates a vector of length 50 (50 elements) from 1 to 25 equally spaced

(vect11 <- 0:30)
(length(vect11)) # function length before a vector object outputs vector length 
```

#### Calculations using vectors

```{r}
vect4 * 2 #each element in the vector is multiplied by 2

vect4 - 10 #10 is subtracted from each of the elements

vect4 / 5 #each element of vect5 is divided by 4

vect4^5 #each element is raised to the power of 4

sqrt(vect4) #outputs square root of each element of vector4

sqrt(vect4[31]) #outputs sqrt of element at position 31 of vect4
```

#### Retrieving vector elements
Square brackets `[]` are used to index the position of elements in any data structure, vectors included. Lowest index of an element is 1 and not 0.

```{r}
(vect4 <- 0:61)
length(vect4)
vect4[25] #retrieves element at position 25 in vector4
vect4[26] #retrieves element at position 26 in vector 4
vect4[62] #retrieves element at position 26 in vector 4
```

#### Converting vectors from one type to another
```{r}
(data5 <- c(1, 2, 3, "Secondary", "University"))
class(data5) #character vector

(x <-as.numeric(data5)) #converting to numeric
class(x)

(data6 <- c(1, 2, 3))
class(data6) #numeric vector

(y <- as.character(data6)) #converting to character
class(y)
```

### Matrices
These are similar to vectors however they are bi dimensional. The data is organised into rows and columns. Columns are clearly defined. Is combination of multiple vectors. Use the function `matrix()` to assign an object a matrix.
Numeric matrices can be used to quickly carry out summaries. By default, matrix arranges the elements by the columns, that is the elements increasing from column to column. This can be changed using the `byrow=TRUE` argument. Matrices are homogeneous.

#### Creating a matrix

```{r creatmtx}
(mtx1 <- matrix(c(1:9), nrow=3, ncol=3))

#change order of elements
(mtx2 <- matrix(c(1:9), nrow=3, ncol=3, byrow = TRUE))
```

#### Retreiving elements of a matrix
```{r matrixs}
mtx1

mtx1[2,2] #pulls the element at row2 in column2

mtx1[,3] #pulls all elements in column 3

mtx1[3,] #pulls all elements in row3

mtx1[7] #pulls the element at index 7

mtx2[c(1,3),] #pulls everything in rows 1 and 3

mtx2[c(1,3)] #returns values at index 1 and 7

mtx2[,c(2,3)] #returns column 2 and 3

mtx2[,1:3] #returns column 2 to 3
```

#### Calculations using matrices
```{r calcmtx}
#summary
summary(mtx1)
summary(mtx2)

#Further calculations
(mtx3 <- matrix(c(seq(from= -98, to= 100, by=2)), nrow=10, ncol=10))
apply(mtx3,2,mean) #calculating mean of each column, 2 means columns

apply(mtx3,1,mean) #calculating mean of each row, 1 means rows
apply(mtx3,1,sum) #calculating sum of each row, 1 means rows

apply (mtx3,1,function(x)mean(x[x>0])) #mean of rows of only positive numbers

```

### Data Frames
These are two dimensional collection of different data types. The data are organised in unlimited number of rows and columns with each row representing observations and each column representing variables.They are created using the function `data.frame()` to assign an object as a data frame. Each of the columns is vectorised and stores only one type of data. Most of the data collected and used are organised in data frames. To reference a column, use `$` symbol and `[]` for indexing the elements.

#### Creating a data frame

```{r}
(df1 <- data.frame(col1 = c(1, 2, 3, 4, 5,7),
                  col2 = c("ann", "john", "erick", "james", "june", "Jasper")))

class(df1)
typeof(df1)
str(df1)
```

#### Create data frames from different types of vectors
```{r}
(fname <- c("Ken", "Cate", "Amalia", "Jimmy", "dorcas")) #character vector
(score <- c(34, 67, 90, 200, 20)) #numeric vector
(fcolour = factor(c("blue", "green", "red", "pink", "green"))) #factor vector
(gender2 = factor(c("M", "F", "F", "M", "F"))) #factor vector
(truefalse = c(TRUE, TRUE, FALSE, TRUE, FALSE))#logical vector

#Combine above to form data frame
(df = data.frame(fname, score, fcolour, gender, truefalse))

str(df)
```

#### Retreiving elements of a data frame

```{r}
df1[2, 2] #outputs element at row2, column2 from data frame 1=john
df1[6, 2]  #outputs element at row6, column 2 =jasper

df1$col2 #outputs all elements of column2 , $reference the column 

df1$col2[5] #outputs the element at position 5 of column 2=june

df1[3, ] #display all elements in row3 -row number
df1

df1[df1$col3 < 12, ] #display rows with col3 less than 12
df1[df1$col5 == 1, ] #display rows with col5 equals 1
df1[df1$col2 == "john", ] #display rows with col2 containing john
df1[df1$col6 > 1, ] #display rows with col6 greater than 1
```

#### Calculations using elements of a data frame

```{r}
(df1$col3 <- df1$col1*4) #multiplies each element in column 1 by 4 and assigns them to column 3 of the same data frame

(df1$col4 <- df1$col1/3) #divides each element of col1 by 3 and outputs in col4

df1[, "col4"] #display all elements in col4-col name
df1[, 4] #display all elements in col4-col number

(df1$col5 <- round(df1$col4, 3)) #rounds off col4 to 4 d.p

(df1$col6 <- signif(df1$col4, 3))

```

#### Combining data frames

```{r}
(df2 <- data.frame(
  var1 = 1:15,
  var2 = 16:30
))

(df3 <- data.frame(
  var3 = 1:15,
  var4 = 16:30
))


(df4 <- data.frame(
  var1 = 16:30,
  var2 = 31:45
))

(df5 <- cbind(df2, df3)) #join two data frames row wise. Must have equal number of rows
(df6 <- rbind(df2, df4)) #appends a data frame to another, col names must be similar

(df2 <- data.frame(
  var1 = 1:15,
  var2 = 16:30
))

(df2$var3 <- paste(df2$var1, df2$var2)) #concatenates var1 and var2 with space in between
(df2$var4 <- paste0(df2$var1, "-", df2$var2)) #concatenates var1 and var2 with a hyphen and no space between
```

### Lists
Is one dimensional structures that can hold different types. It's heterogeneous
Use the function `list()` to create a list. It's like a bucket of different items put together. Individual elements of a list can be extracted by referencing the name of object in the list using `$` and by position `x[[i]]`.
You can embed vectors, matrices and data frames all in a list.

#### Creating lists
```{r}
(list0 <- c("look", 1, TRUE, FALSE)) #holds anything parsed to it

(list1 <- list(
  var1 = 1:20,
  var2 = rep(c("A", "B"), 10)
))

```

#### Retreiving elements of a list
```{r}
list1$var2[6]

list1[1]

list1[[1]][6]
list1$var1[6]

(list2 <- list(
  list1 = list(var1 = 1:20,
               var2 = rep(c("A", "B"), 10)),
  var3 = rep(c("A", "B"), 10)
))


class(list2)
typeof(list2)
str(list2)  

list2$list1$var1[1]
list2[[1]][1]

list2$list1$var2[1]
list2[[2]][1]


(list_all <- list(
  df1 = list(var1 = 1:20,
             var2 = rep(c("A", "B"), 10)),
  df2 = list(var3 = 1:20,
             var4 = rep(c("A", "B"), 10))
))

```

#### Convert list to dataframe

```{r}
#(list_all1 <- do.call(list_all)) #investigate
```


### Arrays
Are multidimensional. Can be more than 1 dimensions and as many as user would wish. Can be considered super set of matrices. Are homogeneous. Can be created by combining matrices. You cannot view an array in one go but can access the levels of the arrays (each of the matrices combined). Useful when stacking matrices on top of each other.
Arrays cannot be used to perform calculations but to store multiple matrices.

#### Creating an array
```{r}
v1 =c(1:12)
(m1 = matrix(v1, nrow=3, ncol=4, byrow=TRUE))

v2 = c(25:36)
(m2 = matrix(v2, nrow=3, ncol=4, byrow=TRUE))

(A = array(c(m1,m2), dim=c(3,4,2)))# 3=rows, 4=columns, 2=number of matrices/levels in the array
```

#### Retreiving levels of an array
```{r}
A[,,1] #level 1
A[,,2] #level 2
```

#### Defining array rows and columns
```{r}
names <- c("jimmy","july", "feb") #row names
info <- c("Height", "Weight", "Hair colour", "skin colour") #column names
(A = array(c(m1,m2), dim=c(3,4,2), dimnames= list(names, info)))
```

#### Retreiving values of an array from different levels
```{r}
A[1,,1] #returns first row of first level
A[1,,2] #returns first row of second level
```

## Functions and Loops
Functions are group of instructions that help users perform/complete a specific task. They can be as complex as an `R package` (collection of functions) or as simple as the assign symbol `(<-)`. Most functions have so far written and can be automatically accessed in R while others need to be access via installing R packages but sometimes one needs a customised function to complete a task. Curly brackets `{}` are used to mark the beginning  and end of instructions to be performed. The `return()` function is used to explicitly instruct R to print the output/result of the last instruction supplied in a function.

Functions have arguments enclosed in a bracket. The arguments are the inputs that the function would use during processing. There must always be an output in a function sometimes produced by the return function.
Functions in R programming are like macros in SAS or STATA programming.

### Available Functions
The functions below (not all)are already available can be used to perform specific tasks as desired.

```{r}
v <- c(2,3,45) #assignment symbol for assigning instructions as an object

c(2,3,"x") # c is function that collects together everything in the brackets

sum(2,3,4) #sum functions returns the sum total of items in the bracket

plot(cars) #returns a plot of dist vs speed from cars dataset

hist(cars$speed) #returns a histogram

attach(cars) # attach function when used before a dataset name, helps user to avoid attaching dataset name to a variable name.

hist(dist) #works without attaching dataset name to the variable

#View(cars) #enables one view the dataset

head(cars) #returns first six observations

tail(cars) #returns last six observations

#data() #lists all datasets available in a package

summary(cars) #returns min,1st QTR, median, mean, 3rd QTR, max of all numeric variables in the dataset.

length(cars) #returns total number of variables in the dataset
length(speed) #returns total number of observations for the variable

unique(dist) #returns all unique entries in a variable

cars[10:15,1:2] #[]subsets data to select observations from row 10 to 15 for all the columns
#function(x) #the function itself is a function that instructs R to treat an object as a function for performing that task.

#%>% #pipe symbol that represents "THEN" and moves processing moves to the next line of instruction. Is handy when you want to avoid nesting of functions.

```

### Creating a function
We write functions to help us to write code in easy to understand chunks. Also, it helps to write reusable code avoid repetitive code lines. A functional function must have certain parameters: A function `name` which is the assigned to the function `function()`, `inputs` or arguments supplied inside the brackets (function(x,y,z)), the curly brackets `{}` to enclose the instructions to be performed, `instructions` to performed supplied inside the curly brackets, then finally the `return()` to output the result of the last instruction performed. Below is a general structure of a function.

functionname <- function(inputs){
  
output_value <- do_something(input)
    
return(output_value)

}

The curly brackets indicate that the instructions inside them are a group hence must be processed together.

```{r}
#Function to calculate the volume of a box
    # inputs: length=23, width= 20, height=10

boxvolume <- function(length, width, height){
  area_of_crosssection <- length*height
  volume <- area_of_crosssection * width
  return(volume)
}

#calling the function
boxvolume(23, 20,10)

#store the results for use later
bvolume <- boxvolume(23, 20,10)
```

### For Loops
For Loops are used for repeating things in R. They are fundamental structures for performing repetitions in  programming. They work by performing the same actions for each item in a list of items like a vector. The key word is `for` which means for each item in the list of items supplied. Basic structure looks like this:

for(item in vector_of_items){

do_something(item)

}

You need to first create a vector of items to loop over.
There are two variants of the For loops. By value and by index.

#### For Loops by value
For loop by value uses the value of a vector to loop over and over until all the elements of the vector have been processed.

```{r}
#A For Loop to calculate masses from each volume of the box

volumes <- c(1.6, 3, 8, 9)
for(vol in volumes){
  mass <- 2.65*volumes^0.95
  print(mass)
} #not sure why four rows are created
```

#### For Loops by index
For loops by index loops over a vector using the index of each element typically starting at 1. The integers (indices) are used to access the values we need from one or more vectors at the position indicated by the index.

```{r}
volumes <- c(1.6, 3, 8, 9)

for(i in 1:length(volumes)){
  mass <- 2.65*volumes[i]^0.95
  print(mass)
}
```

Looping by index helps in storing the results of the loop for use later unlike looping by values. To store the values of the loop, create an empty object of the same length as the vector of items.

```{r}
volumes <- c(1.6, 3, 8, 9)
masses <- vector(length=length(volumes), mode="numeric") #vector function creates an empty vector with argument length, mode option changes the type of vector

for(i in 1:length(volumes)){
  mass <- 2.65*volumes[i]^0.95
  masses[i] <- mass #stores the outputs as it loops
}

masses 
```

Looping by index also allows users to loop over multiple objects at the same time. For example if you want to change the constant 2.65 and exponent 0.95 depending on the circumstance

```{r}
volumes <- c(1.6, 3, 8, 9)
cons <- c(2.65, 3.4, 6.1, 4.3)
expo <- c(0.95, 1.1, 1.5, 2.5)

masses <- vector(length=length(volumes), mode="numeric") #vector function creates an empty vector with argument length, mode option changes the type of vector

for(i in 1:length(volumes)){
  mass <- cons[i]*volumes[i]^expo[i]
  masses[i] <- mass #stores the outputs as it loops
}

masses 
```

#### Looping over files
The code below is commented out as there are no files yet in the training folder directory.

```{r}
#data_files <- list.files(pattern="") #get list of files you want
#results <- vector(length=length(data_files), mode="integer") #create an empty vector of length equal to number of files

#for(i in 1:length(data_files)){
#  data <- read.csv(data_file[i]) #read the csv file (if they are csv files)
#  count <- nrow(data) #count observations if this is what your want
#  results[1] <- count #store the counts for each data files
#}
```

### While Loops
While loops allows you to the code to repeat while a condition is true and stops when the condition is no longer true.

```{r}
#What will be the amount at the end of a year after monthly deposit of 1000 and interest rate of 8% on monthly balance

Whileloop <- function(){
  
  acc <- c(0)
  Month = 1
  
  while (Month<=12){
    
    acc <- (acc*1.08)+1000 #replace the first vector after each calculation
    print(acc)
    
    Month <- Month + 1 #replace the counter with the next increment
    #print(counter)
  }
}

Whileloop()

#Above is similar to For loop but no need of adding the iterator
Forloop <- function(){
  
  acc=0
  
  for (Month in 1:12){
    acc <- (acc * 1.08) +1000
    
    print(acc)
    #print(Month)
  }
    
}

Forloop()

#How do I get just the last value?
```

### Looping Using the if Statements
You can combine loops and functions by putting most of the codes we want to run in a function the calling one more functions each time through the loop. As an example we can use a non-vectorised function to estimate the mass. This type of loop uses the `if` and `else` functions.

If statements allows users to write functions that outputs a result if a condition is true.

```{r}
#Example 1 : Combined if statement and For Loop

est_mass <- function(volume){
  if (volume>5){
    mass <- 2.65*volume^0.95
  } else {
    mass <- NA
  }
  return(mass)
}

volumes <- c(1.6,3,8,4,9)
masses <- vector(length=length(volumes), mode="numeric")

for(i in 1:length(volumes)){
  mass <- est_mass(volumes[i])
  masses[i] <- mass
}

masses

(mass_sapply <- sapply(volumes, est_mass)) # this code is similar to the above for loop

```


```{r}
#Example 2
#Test what to wear depending on  temperature of the day.

Hoodie <- function(temperature){
  
  if (temperature <25){
    print("Wear hoodie")
  } 
  else if (temperature == 25){
    print("Wear vest")
  } 
  else {
    print("Wear nothing")
  }
}

Hoodie(26)
```

## R Packages
R provides a wide range of packages that enable users to perform specific tasks efficiently in R environment. An R package is a collection of functions and can be downloaded from CRAN (Comprehensive R Archive Network). CRAN is the repository for R packages. Some packages already exist in the R environment but some of them must be installed first before they can used. Some packages exists together in an ecosystem of packages for specific tasks. For example the tydyverse package is an ecosystem of packages that enable users clean data, produce graphs and charts, transform data, import data and handle dates

### Installing and loading a package
Installing an R package is very simple. The function `install.packages("packagename")`is used. This makes the package available for access to RStudio but cannot be used yet. One has to load the package into R environment to be able use it and this is done using the function `library(packagename)`. Some packages can exist in more than one ecosystem and one might wish to explicitly indicate the ecosystem from which the package is from by using using double colons as `ecosystemname::packagename`. One has to have an internet connection to be able install packages.

Below are examples of packages commonly used.

`install.packages("magrittr")`

`install.packages("table1")` 

`install.packages("arsenal")`

`install.packages("tinytex")` 

`install.packages("readxl")` 

`install.packages("haven")`

`install.packages("gmodels")`

```{r}

#library(tidyverse) #en ecosystem (collection) of packages like the dplyr,ggplot2, lubridate, tidyr, tibble, stringr, readr, for data cleaning, manipulation, transformation, visualization

#library(magrittr)   # package for piping i.e %>% piping is handy when you want to avoid nesting of functions.

#library(table1) #for descriptive data analysis in HTML

#library(arsenal) #for running proc compare on two datasets and other complex tasks

#library(tinytex) #contains functions for compiling LaTex documents

#library(readxl) #for importing xls dataset

#library(haven) #for importing and exporting SPSS, STATA and SAS files

#library(gmodels) #package for model fitting

#library(knitr) #package for dynamic reporting in R

```

## Data Wrangling
Data wrangling includes a series of processes that involves tidying data  that ends up with suitable datasets for analysis. This starts from reading in data from where it's stored and having a clean data after all the processes. The tidyverse ecosystem of packages has packages of functions that support data wrangling. Package for reading in data `readr`, for generating visuals `ggplot2`, for tidying data `tidyr`, for formatting dates `lubridate`, for modifying a data frame for efficient reading `tibble` and for transforming data `dplyr`. These packages have functions inside them that support what they are intended for.

### Setting work directory
You need to know your working directory before reading in data into the R environment. To know the current working directory use the function `getwd()`. If the directory is different from where you need to save your work, you need to change/set your work directory using the function `setwd()` copy the path to where you want to save work the paste in the setwd() function all wrapped in `""`. Change the backslashes `\` to forward slashes `/` in the path directory.

Use `dir()` to view the files in the directory.

```{r path}
#Knowing your directory
getwd()
dir() #get list of files in the 

#You can save the list of the files in a R object in case you would want to access any of the items in the list.

(file <-dir())

list.files() #this is an alternative to dir()

```

### Reading and writing datasets
The readr package of the tidyverse has functions that enables R users to read and write/save files of different formats.

`CSV files` are text files that have data/values separated by commas. The `read.csv(csvfilename)` functions from readr reads in the csv file and converts it to a data frame for use in R. The `write.csv` function saves he csv file into the working directory. Run `?read.csv` or `?write.csv` to check for the options allowed for these functions.


`Excel` files can include `xls` or `xlsx`. These are files with data arrange in rows and columns. R has the package `readxl` that allows user to read in these files. Use the function `read_excel(excelname)` to read the files. `read_xls()` and `read_xlsx()` also exists for reading excel files. Run `?read_excel` to learn about the options that the function can take.

```{r}
#(df <- read_excel("Dataset/STIData.xls")) #practice data for this training
#View(df) #you can view the data to see all the variables and observations.
#R will always rename variables that have similar names as has happened with the sex variable. 
```

`Text` files are files with values separated by various characters; comma, space, pipe, or semi colon. The `read.table()` function is inbuilt R for reading such files. I requires no package installation. The function takes a number of options. The function is similar to read.csv in reading in csv files however, one has to specify the value separator. Run `?read_table` to learn about the options that the function can take.

```{r}
#(df <- read.table("Dataset/filename.txt", ""))
```

R also support reading data from other soft ware eg `SAS, STATA or SPSS` files. The package `haven` has a number of functions has allows one to reading in file formats from other software.

You can also read in html tables, tables on web pages. The function `readHTMLTable()` allows you to download a webpage table and analyse. 

To connect to databases to read tables directly from a database, one needs the function `dbConnect(dbname(), user="username", password="password", dbname="databasename", host="hos")`

### Exploring data
Before working with your data you need to understand the data. This entails knowing number of your variables and their characteristics, the dimensions of the data, number of observations and the structure of your data. The tidyverse package has packages and functions that helps you explore and clean data. 

Exploring the `STIData`, a real study conducted in Kenya. The `starwars` data is also a good one to explore.

Below are some of the functions you can use to explore your data.

```{r}
library(tidyverse) #first loading the package you will need
#library(readxl)  #load the package for reading the kind of the format
#(df <- read_excel("Dataset/STIData.xls")) #read in your data
#View(df) #view your data
#dim(df) #knowing the dimensions of your data, number of observations and variables
#str(df) #displays the structure of the data
#attach(df) #avoid the need of attaching the name of dataset to varnames
#head(df) #have a glimbse of the first 6 variables
#tail(df) #have a glimbse of the last 6 variables
#df[10:30] #have a look at observations from rows 10 to 30
#names(df) #lists variable names
#length(df) #variables in the dataset
#length(DurationOfillness) #total observations in the variable
#class(DurationOfillness) #types of variable
#unique(Sex...35) #unique entries in the variable
#table(A5MaritalStatus) #table summary of variable, but doesn't count missing values
##View(sort(table(IdNumber), decreasing=TRUE)) #wrapping functions
#summary(Weight) #measure of spread
#boxplot(Weight) #gives a quick view of distribution, box and whiskers
#hist(Weight) #view distribution of any numeric variable, good data should be normally distributed

```

### Cleaning data
#### Handling Missing data
#### Handling dates and times
#### Handling strings
#### Handling dupicates
#### Handling factor variables
#### Restructuring data

## Exploratory Data Analysis
### Univariate data analysis
### Multivariate Data analysis

## Data Visualisation
### Boxplots
### Piecharts
### Histograms
### Barcharts
### Heat maps
### Bubble graphs

## R Shiny
### Building and interactive dashboard
### Developing an RShiny Web App

# ADVANCED PROGRAMMING
